Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-08 paper/paper.tex
Elements of Category Theory

\documentclass{llncs}
% \documentclass[llncs, a4paper, UKenglish, cleveref, autoref, thm-restate]{article}
\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \addbibresource{references.bib}
% \usepackage{amsmath, amssymb, amsfonts, amsthm, stmaryrd}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{enumitem}
% \usepackage{mathpartir}
% \usepackage{bussproofs}
\usepackage{xparse}
\usepackage[usenames, dvipsnames]{xcolor}
\usepackage{lipsum}
\usepackage{xargs}
\usepackage{hyperref}
\usepackage{tikz-cd}
\usepackage{todonotes}
\usepackage{url}
\usepackage{xspace}
\usepackage{rotating}
\usepackage{quiver}
\usepackage{minted}
\usepackage{newunicodechar}
\usepackage{microtype}
% \addbibresource{references.bib}
\usepackage{array}   % for \newcolumntype macro
\newcolumntype{C}{>{$}c<{$}} % math-mode version of "l" column type
\newcolumntype{L}{>{$}l<{$}} % math-mode version of "l" column type

@@

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
%\newcommand{\coq}[1]{\mintinline{coq}{#1}}
\newmintinline[coq]{coq}{fontsize=\small}
\newmintinline[ocaml]{OCaml}{fontsize=\small}
\newminted{coq}{fontsize=\small}
\newminted{ocaml}{fontsize=\small}

@@

% % % \subsubsection{Histomorphism}
% % % The definition of $\lambda$ is:
% % % \begin{align*}
% % %   & \lambda : GG_{\infty} A \to G_{\infty} GA\\
% % %   & \lambda = \fmap{(\fmap{\head})}\comp \anacofree{\fmap{\tail}}
% % % \end{align*}
% % % For example, for $GX = 1 + X$ we have to define
% % % \begin{align*}
% % %   & \lambda : 1 + G_{\infty} A \to G_{\infty} (1 + A)\\
% % %   & \lambda(\inl) = \pair{\inl}{\lambda(\inl)}\\
% % %   & \lambda(\inr(table)) = \pair{\inr (\head(table))}{\lambda (\tail (table))}
% % % \end{align*}
% % % % \begin{align*}
% % % %   & \operator{lookup} :: \Nat \to G_{\infty} \R \to \operator{Maybe}\;\R\\
% % % %   & \operator{lookup}\; \Zero       (\Cons{a}{\text{\textunderscore}})     = \Just{a}\\
% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Zero})                      = \Nothing\\
% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Succ table})                = \operator{lookup}\; n  table
% % % % \end{align*}
% % % % Defining the algebra
% % % % \begin{align*}
% % % %   & \operator{alg} :: D (G_{\infty} \R) \to \R\\
% % % %   & \operator{alg}\; Zero           = 0\\
% % % %   & \operator{alg}\; (Succ{table}) = maximum_{0} [v + \operator{lookup}\; table (i - w) | (w,v) \leftarrow wvs, w \le i]
% % % % \end{align*}
\section{Recursion Schemes}
\subsection{Functors}
Functors are functions \coq{F : Type -> Type} which additionally have a
map
\begin{coqcode}
fmap : forall A B, (A -> B) -> F A -> F B
\end{coqcode}
witnessing the idea that a functor represents a container for abstract data that
can be manipulated without changing the outer structure. For example, the type
\coq{List : Type -> Type} is a functor and \coq{List A} is the type of lists
over elements of type \coq{A} with the obvious \coq{fmap} function recursively
traversing a list and applying the function \coq{A -> B} for each element.
Additionally, the type of lists \coq{List A} arises as the least fixed-point of
the functor \coq{F X = unit + (A * X) }.

In general not every functor has a fixed-point, and it is not possible to build
the fixed-point of a functor such as \coq{F} in Coq due to not being strictly
positive. Due to this, we are going to restrict to \emph{polynomial functors}
as represented by containers.
\subsection{Algebras and Coalgebras}
\subsection{Hylomorphisms}
Recursion schemes provide an abstract way to consume and generate data. We now
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together these
notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:rec-coalgebras}).

\subsection{Algebras and Catamorphisms}
\label{sec:algebras} An algebra is a set \coq{A} (the \emph{carrier} of the
algebra) together with some operations on it sometimes subject to some
equational laws. For example, given such a type \coq{A} we can define the monoid
operations as \coq{u : unit -> A} for unit and \coq{m : A * A -> A} for
multiplication. Notice that \coq{F X = unit + X * X} is a functor on \coq{X}
which means we can equivalently describe these two maps as a single one of type
\coq{F X -> X} which we call the algebra for the functor \coq{F} satisfying the
unit and associative laws of the monoid. In this work we will be using algebras
to describe the operations of a data type and so we will not be needing
additional equations on these operations.

@@

>>>>

\documentclass{llncs}
% \documentclass[llncs, a4paper, UKenglish, cleveref, autoref, thm-restate]{article}
\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
% \addbibresource{references.bib}
\pagestyle{plain}

\usepackage{amsmath, mathpartir}
%, amssymb, amsfonts, stmaryrd}
% \usepackage{mathpartir}
% \usepackage{bussproofs}

\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{xparse}
\usepackage[usenames, dvipsnames]{xcolor}
\usepackage{lipsum}
\usepackage{xargs}
\usepackage{hyperref}
\usepackage{tikz-cd}
\usepackage{todonotes}
\usepackage{url}
\usepackage{xspace}
\usepackage{rotating}
\usepackage{quiver}
\usepackage{minted}
\usepackage{newunicodechar}
\usepackage{microtype}
% \addbibresource{references.bib}
\usepackage{array}   % for \newcolumntype macro
\newcolumntype{C}{>{$}c<{$}} % math-mode version of "l" column type
\newcolumntype{L}{>{$}l<{$}} % math-mode version of "l" column type

@@

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
%\newcommand{\coq}[1]{\mintinline{coq}{#1}}

\newcommand{\Hom}{\text{Hom}}
\newcommand{\Obj}{\text{Obj}}
\newcommand{\Arr}{\text{Arr}}

\newcommand{\List}[1]{\operator{List}(#1)}
\newcommand{\nil}{\operator{nil}}
\newcommand{\cons}{\operator{cons}}

\newmintinline[coq]{coq}{fontsize=\small}
\newmintinline[ocaml]{OCaml}{fontsize=\small}
\newminted{coq}{fontsize=\small}
\newminted{ocaml}{fontsize=\small}

@@

% % % \subsubsection{Histomorphism}
% % % The definition of $\lambda$ is:
% % % \begin{align*}
% % %   & \lambda : GG_{\infty} A \to G_{\infty} GA\\
% % %   & \lambda = \fmap{(\fmap{\head})}\comp \anacofree{\fmap{\tail}}
% % % \end{align*}
% % % For example, for $GX = 1 + X$ we have to define
% % % \begin{align*}
% % %   & \lambda : 1 + G_{\infty} A \to G_{\infty} (1 + A)\\
% % %   & \lambda(\inl) = \pair{\inl}{\lambda(\inl)}\\
% % %   & \lambda(\inr(table)) = \pair{\inr (\head(table))}{\lambda (\tail (table))}
% % % \end{align*}
% % % % \begin{align*}
% % % %   & \operator{lookup} :: \Nat \to G_{\infty} \R \to \operator{Maybe}\;\R\\
% % % %   & \operator{lookup}\; \Zero       (\Cons{a}{\text{\textunderscore}})     = \Just{a}\\
% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Zero})                      = \Nothing\\
% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Succ table})                = \operator{lookup}\; n  table
% % % % \end{align*}
% % % % Defining the algebra
% % % % \begin{align*}
% % % %   & \operator{alg} :: D (G_{\infty} \R) \to \R\\
% % % %   & \operator{alg}\; Zero           = 0\\
% % % %   & \operator{alg}\; (Succ{table}) = maximum_{0} [v + \operator{lookup}\; table (i - w) | (w,v) \leftarrow wvs, w \le i]
            % % % % \end{align*}
\section{Recursion Schemes}
The structure of the a certain data is very similar to the structure of an
algorithm which processes that data. Recursion schemes aim at analyzing these
analogies.

\subsection{Elements of Category Theory}
A \emph{category}, denoted with $\CatC, \CatD, \CatE$,  is a collection of
objects $A, B, C$, denoted by $\Obj(\CatC)$ and a collection of arrows $f, g, h$
between these objects, denoted by $\Arr{\CatC}$, such that there always exists
an identity arrow $id_{A} : A \to A$ for each object $A$ and for two arrows
$A \xrightarrow{f} B$ and $B \xrightarrow{g} C$ there always exists an arrow
$A \xrightarrow{g\circ f} C$ obeying the associativity law. We denote
$\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to $B$\footnote[1]{For
presentation purposes we shall not deal with size issues and assume all the
categories are small.}. The initial object in a category, denoted by $0$, is the
object such that for any other object $A$ there is a unique arrow
$0 \xrightarrow{!} A$.  Dually, the terminal object, denoted by $1$, is the
object such that for any other object $A$ there is a unique arrow
$A \xrightarrow{!} 1$. Initial and terminal object are unique up-to isomorphism.

For example, the category of sets, denoted by $\Set$, is the category where
objects are sets and arrows are functions between sets. The initial object $0$
in $\Set$ is the empty set $\emptyset$ and the terminal object $1$ is any
singleton set. In this section, we are going to assume types are sets, hence
this category can also be viewed as the category of types and programs between
them.

A \emph{functor} $F : \CatC \to \CatD$ is a map between categories mapping both
objects and arrows from one category to another. Hence a functor has two
components, one which maps objects into objects
$F : \Obj(\CatC) \to \Obj(\CatD)$ and one which maps arrows into arrows
$F : \Hom_{\CatC}(A, B) \to \Hom_{\CatD} (FA,FB)$ such that identity and
composition of arrows are preserved:
\begin{mathpar}
  F(id_A) = id_{FA} \and F(g \circ f) = F(g) \circ F(f)
\end{mathpar}
This latter component is also called the \emph{functorial action} and can be
thought of as the $\operator{fmap}$ program in functional programming.

For example, in $\Set$, we can define the set of lists
$\List{A} \iso 1 + A \times \List{A}$ as the set inductively generated by the
constructors $\nil : 1 \to \List{A}$ and
$\cons : A \times \List{A} \to \List{A}$. This type is a functor with functorial
action $\List{f} : \List{A} \to \List{B}$ given by $\List{f}(*) = *$ and
$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
$\List{f}$ is well-defined as it recursively calls on a smaller argument,
however, there is a more general way of proving these type of definitions are
well-defined.

\subsection{Algebras and Catamorphisms}
\label{sec:algebras}
For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(A,\alpha)$
where $A$ is an object of the category called the \emph{carrier} of the algebra
and $\alpha$ is an arrow of type $FA \to A$ called the \emph{structure map}.
The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
objects are $F$-algebras and arrows $f : (A, \alpha) \to (B, \beta)$ in
$F\Alg(\CatC)$ are arrows $f : A \to B$ in $\CatC$ such that they respect the
structure of the algebra, that is $f \circ \alpha = \beta \circ F(f)$.  The
initial object in this category is called the \emph{initial $F$-algebra}, that
is the $F$-algebra which has a unique $F$-algebra homomorphism into any other
$F$-algebra. If $(A, \alpha)$ is initial this implies $A$ is the \emph{least
fixed-point} for the functor $F$, i.e. $A \iso FA$ is the least solution of the
domain equation $X \iso F X$ by Lambek's lemma.

For example, for a set $A$ and a given the functor $F : \Set \to \Set$ mapping
$X \mapsto 1 + A \times X$ an algebra is a set $X$ together with an arrow
$\alpha : 1 + A \times X \to X$.in $\Set$ given such a type
\coq{A} we can define the monoid operations as \coq{u : unit -> A} for unit and
\coq{m : A * A -> A} for multiplication. Notice that \coq{F X = unit + X * X} is
a functor on \coq{X} which means we can equivalently describe these two maps as
a single one of type \coq{F X -> X} which we call the algebra for the functor
\coq{F} satisfying the unit and associative laws of the monoid. In this work we
will be using algebras to describe the operations of a data type and so we will
not be needing additional equations on these operations.

@@

\subsection{Hylomorphisms}
Recursion schemes provide an abstract way to consume and generate data. We now
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together these
notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:rec-coalgebras}).
